---
title: "Case Study: How Does a Bike-Share Navigate Speedy Success?"
author: John Lee
date: < the current date DD/MM/YYYY >
output: html_vignette
---
This work is an output of the capstone project for Google Data Analytics Professional Certificate.
(https://www.coursera.org/professional-certificates/google-data-analytics) This fictional case study packet is based on the Divvy case study, "'Sophisticated, Clear, and Polished’: Divvy and Data Visualization" written by Kevin Hartman. (https://artscience.blog/home/divvy-dataviz-case-study)
Below describes how the bike-sharing user patterns differ in given data set, and provides findings and suggestions based on the result of the analysis.

Scenario:
You are a junior data analyst working in the marketing analyst team at **Cyclistic**, a bike-share company in Chicago. The marketing director believes the company’s future success depends on maximizing the number of annual memberships. Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently.
From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations.

In 2016, Cyclistic launched a bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that are geo-tracked and locked into a network of 692 stations across Chicago.
The bikes can be unlocked from one station and returned to any other station in the system anytime.
Until now, Cyclistic’s marketing strategy relied on building general awareness and appealing to broad consumer segments. One approach that helped make these things possible was the flexibility of its pricing plans: single-ride passes, full-day passes, and annual memberships. Customers who purchase single-ride or full-day passes are referred to as **casual riders (casual)**. Customers who purchase annual memberships are **Cyclistic members (member)**.

Cyclistic’s finance analysts have concluded that annual members are much more profitable than casual riders. Although the pricing flexibility helps Cyclistic attract more customers, marketing director believes that maximizing the number of annual members will be key to future growth.
Rather than creating a marketing campaign that targets all-new customers, she believes there is a very good chance to convert casual riders into members. She notes that casual riders are already aware of the Cyclistic program and have chosen Cyclistic for their mobility needs.

The marketing director set a clear goal: Design marketing strategies aimed at converting casual riders into annual members. In order to do that, however, the marketing analyst team needs to better understand how annual members and casual riders differ, why casual riders would buy a membership, and how digital media could affect their marketing tactics. She and her team are interested in analyzing the Cyclistic historical bike trip data to identify trends. You have assigned to the first question to the answer: How do annual members and casual riders use Cyclistic bikes differently?

Statement of Purpose:
Use the given data set for 12-month period, analyze different patterns and present findings and suggestions along with the result of analysis. R has used for this work.

Data source:
Cyclistic's historical trip data from 2013 has provided for analysis: https://divvy-tripdata.s3.amazonaws.com/index.html
Note: The datasets have a different name because Cyclistic is a fictional company.
The data has been made available or case study by Motivate International Inc. under the license (https://ride.divvybikes.com/data-license-agreement)
In this study, I have downloaded quarterly data in 2019. Each file is provided with ZIP format. each file contains a csv file.

Data pre-processing
this work is following in below steps
1. import libraries for use.
2. import data from downloaded files.
3. check variable names, attributes in datasets to merge them into a single dataframe.
4. clean up data to prepare analysis.

Step 1. Install and import libraries for use.
```{r}
install.packages("tidyverse")
install.packages("lubricate")
install.packages("ggplot2")
install.packages("readr")
install.packages("dplyr")
install.packages("gridExtra")
```
```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readr)
library(dplyr)
library(gridExtra)
```
Step 2. import data from downloaded files.
```{r}
Trips_2019_Q1 <- read_csv("Divvy_Trips_2019_Q1.csv")
Trips_2019_Q2 <- read_csv("Divvy_Trips_2019_Q2.csv")
Trips_2019_Q3 <- read_csv("Divvy_Trips_2019_Q3.csv")
Trips_2019_Q4 <- read_csv("Divvy_Trips_2019_Q4.csv")
```
Quick check for imported dataframe sizes.
- Trips_2019_Q1: 365,069 rows x 12 columns
- Trips_2019_Q2: 1,108,163 rows x 12 columns
- Trips_2019_Q3: 1,640,718 rows x 12 columns
- Trips_2019_Q4: 704,054 rows x 12 columns

Step 3. check for column names to quick review. To merge data frames into one, column names and attributes should consistent each other.
```{r}
colnames(Trips_2019_Q1)
colnames(Trips_2019_Q2)
colnames(Trips_2019_Q3)
colnames(Trips_2019_Q4)
```
Except `Trips_2019_Q2`, column names in other dataframes are identical as below:
 [1] "trip_id"           "start_time"        "end_time"          "bikeid"            "tripduration"
 [6] "from_station_id"   "from_station_name" "to_station_id"     "to_station_name"   "usertype"
[11] "gender"            "birthyear"

While `Trips_2019_Q2` shows different column names, but the attributes look same.
 [1] "01 - Rental Details Rental ID"                    "01 - Rental Details Local Start Time"
 [3] "01 - Rental Details Local End Time"               "01 - Rental Details Bike ID"
 [5] "01 - Rental Details Duration In Seconds Uncapped" "03 - Rental Start Station ID"
 [7] "03 - Rental Start Station Name"                   "02 - Rental End Station ID"
 [9] "02 - Rental End Station Name"                     "User Type"
[11] "Member Gender"                                    "05 - Member Details Member Birthday Year"

Rename the column names to match to the others.
```{r}
Trips_2019_Q2 <- rename(Trips_2019_Q2,
                        trip_id = "01 - Rental Details Rental ID",
                        start_time = "01 - Rental Details Local Start Time",
                        end_time = "01 - Rental Details Local End Time",
                        bikeid = "01 - Rental Details Bike ID",
                        tripduration = "01 - Rental Details Duration In Seconds Uncapped",
                        from_station_id = "03 - Rental Start Station ID",
                        from_station_name = "03 - Rental Start Station Name",
                        to_station_id = "02 - Rental End Station ID",
                        to_station_name = "02 - Rental End Station Name",
                        usertype= "User Type",
                        gender = "Member Gender",
                        birthyear = "05 - Member Details Member Birthday Year"
                        )
```
Inspect the data types for each column. If all matches, then the identical will return 'true'.
```{r}
df1_types <- sapply(Trips_2019_Q1, class)
df2_types <- sapply(Trips_2019_Q2, class)
df3_types <- sapply(Trips_2019_Q3, class)
df4_types <- sapply(Trips_2019_Q4, class)
identical(df1_types, df2_types)
identical(df1_types, df3_types)
identical(df1_types, df4_types)
```
Combine all quarterly data into single dataframe to show annual data.

```{r}
Trips_2019v0 <- bind_rows(Trips_2019_Q1, Trips_2019_Q2, Trips_2019_Q3, Trips_2019_Q4)
```
Keep this original dataframe for backup. Its size is 3,818,004 rows by 12 columns.

Step 4. Clean up data to prepare analysis.
Drop irrelevant columns ("gender", "birthyear") then save to new data frame
```{r}
Trips_2019v1 <- Trips_2019v0 %>% select(-c(gender, birthyear))
```
Add a new column to show trip duration time in minutes.
```{r}
Trips_2019v1$duration_min <- as.numeric(difftime(Trips_2019v1$end_time, Trips_2019v1$start_time, units = "mins"))

```
According to the values in a new column 'duration_min', Current 'tripduration' column indicates values in seconds. To avoid confusion, change the column name accordingly.
```{r}
names(Trips_2019v1)[names(Trips_2019v1) == 'tripduration'] <- "duration_sec"

```
Check the dataframe Trip_2019v1 using summary() function. the Minimum value in duration_min is -56.37. Since this column has calculated based on the time difference between `end_time` and `start_time`, negative value seems strange. The packet explains that those values are from the bikes are taken out of services. So I will delete those rows from the data.
```{r}
Trips_2019v1 <- subset(Trips_2019v1, duration_min >= 0)
```
Now updated dataframe has 3,817,991 rows, or 13 rows are removed - It won't affect much for the analysis.

(Optional) If your OS language is not set to English, then it would be better to set them to English when extracting dates.
```{r}
Sys.setlocale("LC_TIME", "en_US.UTF-8")
```
Extract data and time information from `start_time` and `end_time`.
I want to get the year, month, start/end date, and weekday information.
```{r}
Trips_2019v1$year <- as.numeric(format(Trips_2019v1$start_time, "%Y"))
#Trips_2019v1$month <- as.numeric(format(Trips_2019v1$start_time, "%m"))
Trips_2019v1$month <- format(Trips_2019v1$start_time, "%b")
Trips_2019v1$days <- weekdays(Trips_2019v1$start_time)
Trips_2019v1$start_date <- as.numeric(format(Trips_2019v1$start_time, "%d"))
Trips_2019v1$end_date <- as.numeric(format(Trips_2019v1$end_time, "%d"))
```
And convert current start/end times to show time data only.
```{r}
Trips_2019v1$start_time <- format(Trips_2019v1$start_time, "%H:%M:%S")
Trips_2019v1$end_time <- format(Trips_2019v1$end_time, "%H:%M:%S")
```




