---
title: "Case Study: How Does a Bike-Share Navigate Speedy Success?"
author: John Lee
date: < the current date DD/MM/YYYY >
output: html_notebook
---
This work is an output of the capstone project for Google Data Analytics Certificate.
Below describes how the bike-sharing user patterns differ in given data set, and provides findings and suggestions based on the result of the analysis.

The [R plugin](https://www.jetbrains.com/help/pycharm/r-plugin-support.html) for IntelliJ-based IDEs provides
handy capabilities to work with the [R Markdown](https://www.jetbrains.com/help/pycharm/r-markdown.html) files.
To [add](https://www.jetbrains.com/help/pycharm/r-markdown.html#add-code-chunk) a new R chunk,
position the caret at any line or the code chunk, then click "+".

library(gridExtra)
```{r}

```
The code chunk appears:
```{r}
print("hello")
```

Type any R code in the chunk, for example:
```{r}
mycars <- within(mtcars, { cyl <- ordered(cyl) })
mycars
```

Now, click the **Run** button on the chunk toolbar to [execute](https://www.jetbrains.com/help/pycharm/r-markdown.html#run-r-code) the chunk code. The result should be placed under the chunk.
Click the **Knit and Open Document** to build and preview an output.
